Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-22 latex/main.tex
Updates from Overleaf

The ARG itself is a useful resource to support future studies of SARS-CoV-2.
Our results show that the ARG captures and unifies important features of SARS-CoV-2 evolution,
which are shown in different studies using different methods, tools, and approaches \cite{Turakhia2022,Bloom2023,Hunt2024}.
The ARG displays remarkable phylogenetic congruence with a state-of-the-art phylogeny built using UShER.
Also, the ARG compactly encodes genetic variation in millions of SARS-CoV-2 genomes,
allowing for convenient and efficient reuse of this valuable dataset
for a variety of evolutionary and genetic analyses beyond those presented here.
% We hope that the ARG will spur on new research areas,
% in which recombination is naturally and fully accounted for,
% and will make an impact comparable to the impact that a recently published human ARG \cite{Wohns2022}
% has made on our understanding of human genetic ancestry and demographic past (e.g., REF).
So far, recent developments in ARG methods have focused almost 
exclusively on applications to human sequencing data. In that space,
the availability of ARGs reconstructed from large datasets has already
had tremendous impact on improving our understanding of human genetic
ancestry and past demography \cite{Wohns2022}. We expect
similar leaps to be enabled through tailoring ARG reconstruction methods
to data from other species and pathogens. Specifically, existing
inference methods can be adapted to use the sc2ts ARG for the 
genealogy-based study of selection \cite{hejase2022deep} or the
geographic spread of the virus through time
\cite{Grundler2025geographic}. This is in addition to simpler
analyses, such as calculating mutation spectra or querying the number
or frequency of mutations at a given locus, which the ARG makes 
significantly more efficient, and ultimately more easily reproducible.
Since the ARG is a condensed representation of the genetic information
contained in the sequencing data, it can also be used to understand,
for instance, how the variability in sequencing efforts across time 
and location affects our ability to reconstruct the origin of 
important lineages and detect new recombinants.
Since reconstructing the ARG is computationally expensive, 
and checking and improving various aspects of its quality is laborious, 
we hope that making this resource openly available will save
significant time and 
effort for other researchers and enable immediate focus on analysing 
the information contained within this data. In-depth analysis of the
ARG is also incredibly useful for identifying important directions for
further research, both regarding methods (where current models do not
correctly capture features of the data) and data processing 
(where gaps in sequencing or data artifacts result in clear
reconstruction errors).

>>>>

The ARG itself is a useful resource to support future studies of SARS-CoV-2.
Our results show that the ARG captures and unifies important features of SARS-CoV-2 evolution,
which are shown in different studies using different methods, tools, and approaches \cite{Turakhia2022,Bloom2023,Hunt2024}.
The ARG displays remarkable phylogenetic congruence with a state-of-the-art phylogeny built using UShER.
Also, the ARG compactly encodes genetic variation in millions of SARS-CoV-2 genomes,
allowing for convenient and efficient reuse of this valuable dataset
for a variety of evolutionary and genetic analyses beyond those presented here.
% We hope that the ARG will spur on new research areas,
% in which recombination is naturally and fully accounted for,
% and will make an impact comparable to the impact that a recently published human ARG \cite{Wohns2022}
% has made on our understanding of human genetic ancestry and demographic past (e.g., REF).
So far, recent developments in ARG methods have focused almost 
exclusively on applications to human sequencing data. In that space,
the availability of ARGs reconstructed from large datasets has already
had tremendous impact on improving our understanding of human genetic
ancestry and past demography \cite{Wohns2022}. We expect
similar leaps to be enabled through tailoring ARG reconstruction methods
to data from other species and pathogens. Specifically, existing
inference methods can be adapted to use the sc2ts ARG for the 
genealogy-based study of selection \cite{hejase2022deep} or the
geographic spread of the virus through time
\cite{Grundler2025geographic}. This is in addition to simpler
analyses, such as calculating mutation spectra or querying the number
or frequency of mutations at a given position, which the ARG makes 
significantly more efficient, and ultimately more easily reproducible.
Since the ARG is a condensed representation of the genetic information
contained in the sequencing data, it can also be used to understand,
for instance, how the variability in sequencing efforts across time 
and location affects our ability to reconstruct the origin of 
important lineages and detect new recombinants.
Since reconstructing the ARG is computationally expensive, 
and checking and improving various aspects of its quality is laborious, 
we hope that making this resource openly available will save
significant time and 
effort for other researchers and enable immediate focus on analysing 
the information contained within this data. In-depth analysis of the
ARG is also incredibly useful for identifying important directions for
further research, both regarding methods (where current models do not
correctly capture features of the data) and data processing 
(where gaps in sequencing or data artifacts result in clear
reconstruction errors).
